A formal approach but practical approach in train regulation

### 2025-06-12 paper/main.tex
modele informel sur les trains

\newcommand{\directions}{{\mathsf{D}}}
\newcommand{\turnouts}{{\mathsf{T}}}
\newcommand{\blocks}{{\mathsf{B}}}
\newcommand{\internalState}{{\mathcal{I}}}
\newcommand{\forward}{{\mathtt{f}}}
\newcommand{\backward}{{\mathtt{b}}}
\newcommand{\sucblock}{{\mathtt{suc}}}

@@

@@

We suppose that the train network can be described as follows. First, we consider given a set of blocks: $\blocks = \{ 1, 2, \ldots, n\}$. Blocks are oriented: there is a $\forward$ end and a $\backward$ one. Let $\directions = \{\forward, \backward\}$ be the two directions.  Next, we suppose given a set $\turnouts = \{ n+1, \ldots, m\}$, the identifiers  of trackworks (switches, crossings, double slip switches, and so on). Trackworks are characterized by an internal state: switches can be in direct or deviate position (next denoted '\direct' and '\deviate'),  a crossing has only one internal position, a double slip switch can be straight through or crossing. Let $\internalState$ denotes the set of internal states. A \emph{network configuration} is given by a function $S: \turnouts \to \internalState$. Let $\Sigma$ denotes the set of all network configurations.

@@

Finally, aside of the tracks, at each end of the blocks, there is a signal whose state is \sigred (RED) or \siggreen (GREEN). In other words, the status of signals is described by a function $F: \blocks \times \directions \to \{ \sigred, \siggreen\}$. Given some signal status $F$, a block $pos$ and a direction $dir$, $F(pos,dir) = \sigred$ means a train in block $pos$ cannot leave it in direction $dir$. The signal status is entirely determined by the network configuration and the locking status due to the regulator. We will be back to it later.  Since we do not model train speed, we do not distinguish signals with speed limits from full-speed ones.

Now, we come to the train model. Each train has an identifier

>>>>

\usepackage{package/formal-grammar}
\usepackage{package/mathpartir}

\newcommand{\directions}{{\mathsf{D}}}
\newcommand{\turnouts}{{\mathsf{\theta}}}
\newcommand{\blocks}{{\mathsf{B}}}
\newcommand{\internalState}{{\mathcal{I}}}
\newcommand{\forward}{{\mathtt{f}}}
\newcommand{\backward}{{\mathtt{b}}}
\newcommand{\sucblock}{{\mathtt{suc}}}

@@

\newcommand{\su}{{\mbox{$\mathtt{startuntil}$}}\xspace}

\newcommand{\trains}{{\mathsf{T}}}

@@

We suppose that the train network can be described as follows. First, we consider given a set of blocks: $\blocks = \{ 1, 2, \ldots, n\}$. Blocks are oriented: there is a $\forward$ end and a $\backward$ one. Let $\directions = \{\forward, \backward\}$ be the two directions.  Next, we suppose given a set $\turnouts = \{ n+1, \ldots, m\}$, the identifiers  of trackworks (switches, crossings, double slip switches, and so on). Trackworks are characterized by an internal state: switches can be in direct or deviate position (next denoted '\direct' and '\deviate'),  a crossing has only one internal position, a double slip switch can be straight through or crossing. In the examples, to simplify the presentation, we use only switches, but our model is not restricted to them.  Let $\internalState$ denotes the set of internal states. A \emph{network configuration} is given by a function $S: \turnouts \to \internalState$. Let $\Sigma$ denotes the set of all network configurations. For two turnouts $\theta = \{9, 10\}$, the function $9 \mapsto \deviate, 10 \mapsto \direct$ states that $9$ is in deviate position while 10 is in direct one. The latter function is also described via an array  $[\deviate, \direct]$ giving states in trackwork order.

@@

Finally, aside of the tracks, at each end of the blocks, there is a signal whose state is \sigred (red) or \siggreen (green). In other words, the status of signals is described by a function $F: \blocks \times \directions \to \{ \sigred, \siggreen\}$. Given some signal status $F$, a block $pos$ and a direction $dir$, $F(pos,dir) = \sigred$ means a train in block $pos$ cannot leave it in direction $dir$. The signal status is entirely determined by the network configuration and the locking status due to the regulator. We will be back to it later.  Since we do not model train speed, we do not distinguish signals with speed limits from full-speed ones.

Now, we come to the train model. Each train has an identifier. In other words, we have a set of trains $\trains = \{1, \ldots, k\}$. Each train has a position on the network. The position is the index of the block it stays in or the last block it was in. Next, a train has a direction in $\directions$. The direction does not mean the train is running---it can be blocked by some red signal---but that its next move will be in that direction. Finally, a train has duties. Duties are given as a sequence of orders following the grammar:
$$  seq ::= \su(dir,pos) $$
with $dir \in \directions$ and $pos \in \blocks$. A duty $\su(dir, pos)$ states that the train direction is set to $dir$ and that the train runs until it reaches $pos$.  The sequence may be empty.
